---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(ggmosaic)
iris <- tibble(iris)
```

# More dealing with data Questions

###  Complete this quiz to evaluate your comprehension of the material
A similar, if not identical quiz should be completed for credit on canvas.

Always [`glimpse()`](https://tibble.tidyverse.org/reference/glimpse.html) first  

```{r glimpse, exercise=TRUE, eval = FALSE}
glimpse(iris)
```





#### Q1 and Q2: `mutate()`, `arrange()`, `select()`   

Use the [`mutate`](https://dplyr.tidyverse.org/reference/mutate.html) and then the [`arrange()`](https://dplyr.tidyverse.org/reference/arrange.html) (in `desc`ending order) functions to find the individual with the greatest `Sepal.Length` relative to (aka divided by) `Petal.Length`, [`select()`](https://dplyr.tidyverse.org/reference/arrange.html) the new column `sepal_to_petal_length` and then answer the question.



```{r mutatearrangeselect, exercise=TRUE, eval = FALSE}
iris %>%
  mutate(sepal_to_petal_length = ___/___) %>%
  ____(desc(____))                        %>%
  select(____)                            %>%
  head(n = 3)                             # let's just see the top 3
```

```{r mutatearrangeselect-solution}
iris %>%
  mutate(sepal_to_petal_length = Sepal.Length/Petal.Length) %>%
  arrange(desc(sepal_to_petal_length)) %>%
  select(sepal_to_petal_length) %>%
  head(n = 3) # let's just see the top 3
```

**Q1**

```{r mutatearrangeselectQ, echo=FALSE}
question_text("What is the largest `sepal_to_petal_length` in the data set (include three digits past the decimal)?",
    allow_retry = T,
    answer("4.833", correct = T),
    correct = random_praise(),
    incorrect = "good try, the answer is 4.833"
)
```


**Q2** 

```{r mutatearrangeselectQ2, echo=FALSE}
question("Above the order was `iris %>% mutate %>% arrange %>% select`. Which other order would still work? (*Hint:* try them and see if it makes sense)",
    answer("`mutate %>% iris    %>% select %>% arrange`"),
    answer("`iris   %>% select  %>% mutate %>% arrange`"),   
    answer("`iris   %>% arrange %>% mutate %>% select`"),  
    answer("`iris   %>% mutate  %>% select %>% arrange`", correct = TRUE),
    answer("order of code doesnt matter"),
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement()
  )
```


#### Q3: `group_by()` and `summarise()`   

Combine the [`group_by()`](https://dplyr.tidyverse.org/reference/group_by.html) and [`summarise()`](https://dplyr.tidyverse.org/reference/summarise.html) functions to find the mean petal width of each iris species 




```{r groupbysummarise, exercise=TRUE, eval = FALSE}

```

```{r groupbysummarise-solution}
iris %>%
  group_by(Species) %>%
  summarise(Mean.Petal.Width = mean(Petal.Width))
```



```{r summarise, echo=FALSE}
question_text("What is the mean Petal Width of the species with the smallest mean Petal Width?",
    allow_retry = T,
    answer("0.246", correct = T),
    correct = random_praise(),
    incorrect = "good try, the answer is 0.246"
)
```


#### Q4 and 5: `ggplot()` 

**Q4:** Plot petal length (on the y) as a function of petal width (on the x), showing species with color

```{r ggplot, exercise=TRUE, eval = FALSE}
ggplot(iris, aes(x = ___, y = ___, ___ = ____))+
  geom_____()
```

```{r ggplot-solution}
ggplot(iris, aes(x = Petal.Width, y = Petal.Length, color = Species))+
  geom_point()
```



**Q5:** Plot sepal width (on the y) as a function of sepal length (on the x), facetting by species

```{r ggplot2, exercise=TRUE, eval = FALSE}

```

```{r ggplot2-solution}
ggplot(iris, aes(x = Sepal.Width, y = Sepal.Length))+
  geom_point()+
  facet_wrap(~Species)
```
