---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(ggmosaic)
iris <- tibble(iris)
ukgas<- data.frame(year = 1960:1986,matrix(c(UKgas), ncol = 4,byrow = TRUE))
names(ukgas) <- c("year","quarter_1", "quarter_2", "quarter_3", "quarter_4")
ukgas <- tibble(ukgas)
knitr::opts_chunk$set(echo = FALSE)
toad_data <- tribble(~individual,~species,~sound,~weight,~height,
        "a", "Bufo spinosus", "chirp", 2.0,	2,
        "b",	"bufo bufo",	"croak",	2.6,	3,
        "c",	"Bufo bufo",	"ribbit",	3.0,	2)
```

# Dealing with data Questions

###  Complete this quiz to evaluate your comprehension of the material
A similar, if not identical quiz should be completed for credit on canvas.

#### Q1: [`glimpse()`](https://tibble.tidyverse.org/reference/glimpse.html)

The `iris` dataset in `R` has floral measurements for fifty individuals each from three of `iris species`. Execute the code below, fix the mistake, and then answer the question which follows.

```{r glimpse, exercise=TRUE, eval = FALSE}
glimps(iris)
```


```{r glimpse-solution}
glimpse(iris)
```



```{r glimpseQ, echo=FALSE}
question("Species is of class character.",
    answer("TRUE"),
    answer("FALSE", correct = TRUE),
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement()
  )
```

  

#### Q2: [`filter()`](https://dplyr.tidyverse.org/reference/filter.html)     

Use the [`filter()`](https://dplyr.tidyverse.org/reference/filter.html) function to find out how many individuals in the dataset have a `Sepal.Length` greater than 7 units, and then answer the question.


```{r filter, exercise=TRUE, eval = FALSE}
filter(iris,  ___ > 7  )
```

```{r filter-solution}
filter(iris,  Sepal.Length > 7  )

## OR 

iris %>%
  filter(Sepal.Length > 7 )
```


```{r filterQ, echo=FALSE}
question_text("How many individuals Sepal.Length's greater than 7 units?",
    allow_retry = T,
    answer("12", correct = T),
    correct = random_praise(),
    incorrect = random_encouragement()
  )
```



#### Q3: Reproducibility 


```{r reproduce, echo=FALSE}
question("It is more important to save the code underlying an analysis than the result of the analysis.",
    answer("TRUE", correct = TRUE),
    answer("FALSE"),
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement()
  )
```

#### Q3-6: [`arrange()`](https://dplyr.tidyverse.org/reference/arrange.html)  again


**Q3:** Find the shortest character in the `starwars` data set [hint, you might want to [`glimpse()`](https://dplyr.tidyverse.org/reference/glimpse.html)  first.]

```{r arrange2, exercise=TRUE, eval = FALSE}
starwars_sortheight <- ____(starwars, ___) 

# Use the to just  show name and height of this new sorted data set
select(starwars_sortheight, ____, ___) 
```

```{r arrange2-solution}
starwars_sortheight <- arrange(starwars, height) 

# Use the to just  show name and height of this new sorted data set
select(starwars_sortheight, name, height) 
```


```{r arrange2Q, echo=FALSE}
question("Who is the shortest star wars character?",
    answer("Yoda", correct = TRUE),
    answer("R2-D2"),
    answer("Luke Skywalker"),
    answer("Yarael Poof"),
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement()
  )
```

**Q4:** 

```{r arrange2Qnochange, echo=FALSE}
question("After executing the code above, who is the first character in the `starwars` data set?",
    answer("Yoda"),
    answer("Luke Skywalker", correct = TRUE),
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement()
  )
```

**Q5:** 

```{r arrange2Qdoublecheck, echo=FALSE}
question("After executing the code above, who is the first character in the `starwars_sortheight` data set?",
    answer("Yoda", correct = TRUE),
    answer("Luke Skywalker"),
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement()
  )
```

**Q6:** 

```{r arrange2Qstillnochange, echo=FALSE}
question("After executing the code above, how many columns are in the ` `starwars_sortheight` dataset?",
    answer("87"),
    answer("14"),
    answer("2", correct = TRUE),
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement()
  )
```


## Q7: Best practices in spreadsheets
In Chapter 2 we discussed best practices in making spreadsheets. Imgine someone messed up, so our `toad_data` looks like this.    

```{r, echo=FALSE}
toad_data
```

If I then filter for `Bufo bufo`

```{r, echo=TRUE}
filter(toad_data, species == "Bufo bufo")
```



```{r spreadsheets, echo=FALSE}
question("Which rule for making good spreadsheets did I break and why does it matter?",
    answer("Values where implied, so I missed relevant context"),
    answer("Difficult to understand column names made it hard to get the species column right"),
    answer("We used color to contain information, which is lost in R"),
    answer("Species names where inconsistent, so we didn't appropriately filter our data", correct = TRUE),
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement()
  )
```

####  OPTIONAL: Tidying data

Let's have a  [`glimpse`](https://tibble.tidyverse.org/reference/glimpse.html) at data for gas consumption in the UK by quarter over years

```{r}
glimpse(ukgas)
```

These data are not tidy! Each observation is a quarter of a year, and the variables are -- `year`, `quarter`, and `gas` let's tidy theme with the [`pivot_longer`](https://tidyr.tidyverse.org/reference/pivot_longer.html) function.

```{r, echo=TRUE}
pivot_longer(data = ukgas, 
             cols = c("quarter_1", "quarter_2", "quarter_3", "quarter_4"),
             names_to = "quarter", 
             names_prefix = "quarter_",
             values_to = "gas") %>%
  head(n = 6)
```



```{r tidying, echo=FALSE}
question("Looking at the code above (and perhaps remembering the video), guess what `names_prefix = 'quarter_'`, does?.",
    answer("Moves quarter_ from values in a cell to the column  name"),
    answer("Removes `quarter_`, from `quarter_1`,`quarter_2`, `quarter_3`, and `quarter_4`.", correct = TRUE),
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement()
  )
```
