# Break  {#break2}   

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(DT)
library(knitr)
library(blogdown)
library(beyonce, warn.conflicts=F, quietly=T)
library(stringr)
library(tweetrmd)
library(emo)
library(tufte)
library(cowplot)
library(lubridate)
library(ggthemes)
library(kableExtra)
library(ggforce)
library(datasauRus)
library(ggridges)
library(randomNames)
library(infer)
library(tiktokrmd)
library(ggridges)
library(colorspace)
options(crayon.enabled = FALSE)   
```

Take this time to check in on your understanding of stats concepts, and R code. 

Be sure you understand  

- Population vs Samples   
- Parameters  vs Estimate   
- The sampling distribution
- Estimation with uncertainty   
- Probability -- probability laws and Bayes' theorem    
- Null hypothesis significance testing -- the null hypothesis -- a p-value etc   
- The binomial distribution  

Please revisit old homework questions the where hard for you, and fill out this [survey](https://forms.gle/qPFGEqZoaiN7rA7t7)

## Reproducible examples

In R it's good to seek help, but great to help people help you.  Watch this video on making reproducible examples. Then make a reproducible example that to the [#reproduciblerhelp](https://biostatsspring2021.slack.com/archives/C01QET9V6UV) slack channel, and try to help one other person!  

```{r, echo=FALSE}
include_url("https://www.youtube.com/embed/G5Nm-GpmrLw")
```


```{r, echo=FALSE}
rm(list = ls())
```
