---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(ggmosaic)
iris <- tibble(iris)
ukgas<- data.frame(year = 1960:1986,matrix(c(UKgas), ncol = 4,byrow = TRUE))
names(ukgas) <- c("year","quarter_1", "quarter_2", "quarter_3", "quarter_4")
ukgas <- tibble(ukgas)
knitr::opts_chunk$set(echo = FALSE)
```

# Final summary qs


###  Complete this quiz to evaluate your comprehension of the material
A similar, if not identical quiz should be completed for credit on canvas.

#### Q1: Calculator   


Using R strictly as a calculator (no functions) find the sample variance of a sample with values -1, 0, and 1, recalling that  $s^2   = \frac{\Sigma(x_i - \overline{x})^2}{n-1}$.

The `iris` dataset in `R` has floral measurements for fifty individuals each from three of `iris species`. Execute the code below, fix the mistake, and then answer the question which follows.

```{r math, exercise=TRUE, eval = FALSE}
mean_vals <- (__ + __  + __) / __
SSx_vals  <- (__ - __)^2 + (__ - __)^2  + (__ - __)^2 
var_vals  <- __ / (__ - 1) 
print(var_vals)
```


```{r math-solution}
mean_vals <- (-1 + 0  + 1) / 3
SSx_vals  <- (-1 - 0)^2 + (0 - 0)^2  + (1 - 0)^2 
var_vals  <- SSx_vals / (3 - 1) 
print(var_vals)
```

#### Q2:  With R



```{r r, exercise=TRUE, eval = FALSE}
# let's make river_length a column in a tibble called river.length
river.length <- tibble(river_length = rivers) 

river.length    %>%
  summarise(mean_length    =  ___(river_length),
            var_length     =  __(__) ,  
            sd_length      =  __(), 
            coefvar_length = 100 * ___ / ___)
```

```{r r-solution}
# let's make river_length a column in a tibble called river.length
river.length <- tibble(river_length = rivers) 

river.length    %>%
  summarise(mean_length    =  mean(river_length),
            var_length     =  var(river_length),  
            sd_length      =  sd(river_length), 
            coefvar_length = 100 * sd_length / mean_length)
```
