---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(ggforce)
iris <- tibble(iris)
```

# Intro to ggplot quiz

###  Complete this quiz to evaluate your comprehension of the material

A similar, if not identical quiz should be completed for credit on canvas.

**Today, we will focus on the** ***iris*** **dataset, which is built-in with R**  

Always [`glimpse()`](https://tibble.tidyverse.org/reference/glimpse.html) first.   

```{r glimpse, exercise=TRUE, eval = FALSE}
glimpse(iris)
```

#### Q1: Distribution of one continuous variable  

```{r onecontQ, echo=FALSE}
question("Which of these are good ways to look at the distribution of a numeric variable. (pick one)",
    answer("A barplot"),
    answer("A boxplot"), 
    answer("A histogram or density plot", correct = TRUE),
    answer("A scatterplot or biplot"), 
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement()
  )
```

Make the plot that best answers the question above for any of the numeric variables in *iris*


```{r onecont, exercise=TRUE, eval = FALSE}
ggplot(iris, aes(___ = ___)) +
  geom_____()
```

```{r onecont-solution}
# potential answers 
ggplot(iris, aes(x = Petal.Length)) +
  geom_histogram()

ggplot(iris, aes(x = Sepal.Width)) +
  geom_density(fill = "pink")
```


#### Q2: One continuous, one categorical   


```{r catcon, echo=FALSE}
question("Which of the following is **A BAD** way to display the relationship between a categorical and continuous variable?",
    answer("barplot", correct = TRUE),
    answer("jitterplot / stripchart"),
    answer("overlapping density plots with appropriate transparency", correct = TRUE), 
    answer("small multiples of histograms"),
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement()
  )
```


Make one (or more) of the good plots suggested above, with any continuous variable and `Species` as a categorical.   

```{r catcontplot, exercise=TRUE, eval = FALSE}
ggplot(iris, aes(___)) +
  ___  # and maybe more
```


```{r catcontplot-solution}
ggplot(iris, aes(x = Petal.Width, fill = Species)) +
  geom_density(alpha = 0.4)  # and maybe more
# OR
ggplot(iris, aes(x = Petal.Width)) +
  geom_histogram()  +
  facet_wrap(~Species, ncol =1)
# OR
ggplot(iris, aes(x = Species, y = Petal.Width)) +
  geom_jitter(width = .25, height = 0, alpha = .5) 
```
  
  

#### Q2: Two continuous, one categorical     

```{r}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species))+
  geom_point()
```

```{r numcat1, echo=FALSE}
question("In the plot above, which species increases its sepal width most strongly with its sepal length?",
    answer("setosa", correct = TRUE),
    answer("Versicolor"),
    answer("virginica"),
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement()
  )
```


```{r numcat2, echo=FALSE}
question("In the plot above, what do you think went in `aes()`",
    answer("x = Sepal.Width,  y = Species,       color = Sepal.Length"),
    answer("x = Sepal.Width,  y = Sepal.Length,  color = Species"),
    answer("x = Species,      y = Sepal.Width,   color = Sepal.Length"),
    answer("x = Species,      y = Sepal.Length,  color = Sepal.Width"),
    answer("x = Sepal.Length, y = Sepal.Width,   color = Species", correct = TRUE),
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement()
  )
```


```{r numcat3, echo=FALSE}
question("In the plot above, what geom was used",
    answer("jitter"),
    answer("point", correct = TRUE),
    answer("line"),
    answer("histogram"),
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement()
  )
```
