---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(ggmosaic)
knitr::opts_chunk$set(echo = FALSE)
```

# Probabilistic math quiz  

## Background

In the last chapter, we imagined there is some disease spreading, and people concerned about the disease take a test with the following behavior:   

- The test correctly reports someone uninfected as uninfected in 0.995 of tests.   
- The test correctly reports someone infected as infected in 0.94 of tests.  



### Simulate disease

Last chapter we tested these people, and find the number that are    

1. healthy and diagnosed as healthy.       
2. healthy and diagnosed as sick.    
2. sick and diagnosed as healthy.   
3. sick and diagnosed as sick.   

Remember 

- That we use the [`case_when`](https://dplyr.tidyverse.org/reference/case_when.html) function to simulate conditional probabilities.    
- The test correctly reports someone uninfected as uninfected in 0.995 of tests.   
- The test correctly reports someone infected as infected in 0.94 of tests.    
- Two in one hundred people have the disease.   
<span style="color:lightgrey;">Note these answers will differ slightly from before because we are running the simulation again. </span>

```{r sim2, exercise=TRUE, eval = FALSE}
n_tests  <- 10000
p_testsick_givensick       <- 0.94
p_testhealthy_givenhealthy <- 0.995
p_sick                      <- 0.02

sickness <- tibble(
  actual_status = sample(c("sick","healthy"), size = n_tests,  replace = TRUE, prob = c(p_sick, 1 - p_sick))) %>%
    group_by(actual_status) %>%
    mutate(test_status = case_when(
      actual_status == "sick" ~ sample(c("sick","healthy"),
            prob = c(___, 1 - ___), 
            replace = TRUE, size = n()),  
      actual_status == "healthy" ~  sample(c("sick","healthy"),
            prob = c(1 - ___, ___), 
            replace = TRUE,  size = n()))) %>%
  ungroup()

### Counting all combinations of outcomes
sickness %>%
  group_by(___, ___) %>%
  summarise(count = n(), .groups = "drop")



#### We will make a plot with the mosaic plot  function in base R. 
# These plots are nice, but hard to make in ggplot. 
# Build off of this code if you ever want to build one. 
# Mosaic plots are discussed in more detail in the textbook
table(sickness) %>% 
  mosaicplot(color = c("firebrick","blue"), 
             main = "", xlab = "actual status", ylab = "test status")
```

```{r sim2-solution}
n_tests  <- 10000
p_testsick_givensick       <- 0.94
p_testhealthy_givenhealthy <- 0.995
p_sick                     <- 0.02

sickness <- tibble(
  actual_status = sample(c("sick","healthy"), size = n_tests,  replace = TRUE, prob = c(p_sick, 1 - p_sick))) %>%
    group_by(actual_status) %>%
    mutate(test_status = case_when(
      actual_status == "sick" ~  sample(c("sick","healthy"),
              prob = c(p_testsick_givensick , 1 - p_testsick_givensick ),
              replace = TRUE, size = n()),  
      actual_status == "healthy" ~ sample(c("sick","healthy"),
              prob = c(1 - p_testhealthy_givenhealthy, p_testhealthy_givenhealthy),
              replace = TRUE, size = n()))) %>%
  ungroup()

### Counting all combinations of outcomes
sickness %>%
  group_by(test_status, actual_status) %>%
  summarise(count = n(), .groups = "drop")

#### We will make a plot with the mosaic plot  function in base R. 
# These plots are nice, but hard to make in ggplot. 
# Build off of this code if you ever want to build one. 
# Mosaic plots are discussed in more detail in the textbook
table(sickness) %>% 
  mosaicplot(color = c("firebrick","blue"), 
             main = "", xlab = "actual status", ylab = "test status")
```

### Exact solutions with math  



#### Oops  

##### Right diagnosis   


**Right diagnosis** Imagine that in one day, 10 healthy people get tested. Find the probability that  all ten get a (correct) negative result (to four digits past the decimal)).   


```{r calculator0, exercise=TRUE, eval=FALSE}
p_testsick_givensick       <- 0.94
p_testhealthy_givenhealthy <- 0.995
p_sick                      <- 0.02

```




<div id="calculator0-hint">
**Hint:** We get correctly diagnosed as healthy by getting one true positive AND another true positive, and another... 

Because we multiply each independent probability, the probability of $n$ consecutive outcomes, each with probability $p$ is  $p^n$.

More generally, the probability of a sequence of outcomes is $\prod_{i=1}^{n}  p_i$, with   $p^n$ as a special case for when the probability of each event in your sequence is the same.  
</div>




```{r exponent, echo = FALSE}
question_text("All ten get a (correct) negative result (to four digits past the decimal))",
 answer("0.9511", correct = TRUE),
 answer( ".9511", correct = TRUE),
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement()
)
```


##### Wrong diagnosis   

**Wrong diagnosis** Imagine that in one day, 10 healthy people get tested. Find the probability that   at lease one  gets an (incorrect) positive result (to four digits past the decimal)   


For math   

```{r calculator1, exercise=TRUE, eval=FALSE}
p_testsick_givensick       <- 0.94
p_testhealthy_givenhealthy <- 0.995
p_sick                      <- 0.02

```


<div id="calculator1-hint">
**Hint:** The probability of not something equals 1 minus the probability of that thing. 

So this should be 1 - the answer above.  
</div>




```{r not, echo = FALSE}
question_text("At lease one  gets an (incorrect) positive result (to four digits past the decimal)",
 answer("0.0489", correct = TRUE),
 answer( ".0489", correct = TRUE),
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement()
)
```

#### Use the mutliplication rule to find 

The proportion of people who are  healthy but diagnosed with the disease  (to four digits past the decimal)"

```{r calculator2, exercise=TRUE, eval=FALSE}
p_testsick_givensick       <- 0.94
p_testhealthy_givenhealthy <- 0.995
p_sick                      <- 0.02

```


<div id="calculator2-hint">
**Hint:** Multiply the probability that someone is healthy by the probability a healthy person gets a false positive.    
</div>



```{r multiplication, echo = FALSE}
question_text("The proportion of people who are  healthy but diagnosed with the disease  (to four digits past the decimal)",
 answer("0.0049", correct = TRUE),
 answer( ".0049", correct = TRUE),
    allow_retry = TRUE,
    correct = random_praise(),
    incorrect = random_encouragement()
)
```

#### Use the law of total probability to find 

Use math to find the proportion of people diagnosed with the disease (to four digits)

```{r calculator3, exercise=TRUE, eval=FALSE}
p_testsick_givensick       <- 0.94
p_testhealthy_givenhealthy <- 0.995
p_sick                      <- 0.02

```


<div id="calculator3-hint">
**Hint:** People can be diagnosed if they are healthy and get a FALSE positive OR if they are sick and get a true positive. Theses are mutually exclusive so we add the probabilities without fear of double counting.
</div>



```{r totalProb, echo = FALSE}
question_text("The proportion of people diagnosed with the disease (to four digits)",
 answer("0.0237", correct = TRUE),
 answer( ".0237", correct = TRUE),
  allow_retry = TRUE,
  correct = random_praise(),
  incorrect = random_encouragement()
)
```


#### Use Baye's theorem  to find 

The proportion of people diagnosed with the disease who actually have it (with four digits past the decimal). That is: number diagnosed and sick / number diagnosed

```{r calculator4, exercise=TRUE, eval=FALSE}
p_testsick_givensick       <- 0.94
p_testhealthy_givenhealthy <- 0.995
p_sick                      <- 0.02

```





```{r bayes, echo = FALSE}
question_text("The proportion of people diagnosed with the disease who actually have it (with four digits past the decimal). That is: number diagnosed and sick / number diagnosed ",
 answer("0.7932", correct = TRUE),
 answer( ".7932", correct = TRUE),
  allow_retry = TRUE,   
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

