---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggmosaic)
library(infer)
library(broom)
knitr::opts_chunk$set(echo = FALSE)

spider <- read_csv("https://whitlockschluter3e.zoology.ubc.ca/Data/chapter09/chap09q07RedbackSpiderCannibalism.csv")
spider_long  <-  spider %>% 
  group_by(secondMaleAcceptance , eatenOrEscape) %>%
  summarise(count = n()) %>%
  ungroup()

spider_wide <- spider_long %>%
  pivot_wider(names_from = "eatenOrEscape", values_from = "count")
```

# Two categorical Quiz

## Background

Sometimes female *Latrodectus hasselti* (redback spiders) eat their mates. Is there anything in it for the males? Maydianne Andrade tested the idea that eating a male might prevent her from re-mating with a second male -- that is if she's too preoccupied eating/digesting her mate that she's not looking to mate again. She observed whether a female accepted a second male after the first male either escaped or was eaten.  


```{r design, echo = FALSE}
question("This study is",
 answer("Observational", correct = TRUE),
 answer("Expiremental"),
 allow_retry = TRUE,   
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

I have these data as   

- All two columns (`secondMaleAcceptance`, `eatenOrEscape`) and 32 rows (one for each 1st male)  in a tibble called `spider`.    
- In long format with three columns (`secondMaleAcceptance`, `eatenOrEscape`, `count`), and four columns in  a tibble called `spider_long`.  
- In wide format as a contingency table with three columns (`secondMaleAcceptance`, `eaten`, `escape`),  with eaten and escaped as count data, in a tibble called `spider_wide`.  (below)   

#### Plot   

**Plot the data** using which data format is easiest for you

```{r plot, exercise=TRUE, eval = FALSE}

```


```{r plot-solution}
ggplot(spider, aes(x = eatenOrEscape, fill = secondMaleAcceptance))+
  geom_bar(position = "dodge")+
  theme(legend.position = "bottom")
```


```{r eyes, echo = FALSE}
question("What trend is apparent from this plot?",
 answer("There is no obvious trend"),
 answer("Following cannibalistic mating females are more likely to accept a second suitor"),
answer("Following cannibalistic mating females are less likely to accept a second suitor", correct = TRUE),
 allow_retry = TRUE,   
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

#### Relative Risk 


```{r, eval=TRUE}
kable(spider_wide) %>% kable_styling()
```




Use the data above to calculate the relative risk of having the second male accepted if you were eaten compared to if you escaped.  

```{r relativerisk, exercise=TRUE, eval = FALSE}
risk_given_eaten    <- __ / (__ + __)
risk_given_escaped  <- __ / (__ + __)
relative_risk       <- __ / __
relative_risk 
```



```{r relativerisk-solution}
risk_given_eaten    <- 3 / (3 + 6)
risk_given_escaped  <- 22 / (22 + 1)
relative_risk       <- risk_given_eaten / risk_given_escaped 
```



```{r rr, echo = FALSE}
question_text("The relative risk of having the second male accepted if you were eaten compared to if you escaped  is(to two places after the decimal)",
 answer("0.35", correct = TRUE),
 answer(".35", correct = TRUE),
 allow_retry = TRUE,   
  correct = random_praise(),
  incorrect = random_encouragement()
)
```


#### Odds ratio   



```{r, eval=TRUE}
kable(spider_wide) %>% kable_styling()
```




Use the data above to calculate the odds ratio  of having the second male accepted if you were eaten compared to if you escaped.  

```{r oddsratio, exercise=TRUE, eval = FALSE}
odds_given_eaten    <- __ / __
odds_given_escaped  <- __ / __ 
odds_ratio         <- __ / __
odds_ratio   
```



```{r oddsratio-solution}
odds_given_eaten    <- 3 /  6
odds_given_escaped  <- 22 / 1 
odds_ratio         <- .5 / 22
odds_ratio  
```



```{r or, echo = FALSE}
question_text("The odds ratio of having the second male accepted if you were eaten compared to if you escaped (to two places after the decimal)",
 answer("0.02", correct = TRUE),
 answer(".02", correct = TRUE),
 allow_retry = TRUE,   
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

## Is this pattern an unexpected outcome of the null   

Hypotheses


```{r hyp, echo = FALSE}
quiz(caption = " Null and alternative hypotheses",
     question("What  is the (two-tailed) null hypothesis",
               answer("There is no association being eaten & your mate rejecting a second male", correct = TRUE),
               answer("Being eaten is not associated with a lower chance of your mate rejecting a second male"),
               answer("There is an association being eaten & your mate rejecting a second male"),
              allow_retry = TRUE,   
              correct = random_praise(),
              incorrect = random_encouragement()),
     question("What  is the (two-tailed) alternative hypothesis",
               answer("There is no association being eaten & your mate rejecting a second male"),
               answer("Being eaten is not associated with a lower chance of your mate rejecting a second male"),
               answer("There is an association being eaten & your mate rejecting a second male", correct = TRUE),
              allow_retry = TRUE,   
              correct = random_praise(),
              incorrect = random_encouragement())
)
```


[Fisher's exact test](https://bookdown.org/ybrandvain/Applied-Biostats/contingency.html#fishers-exact-test-as-an-alternative-to-the-chi2-test.) uses math tricks to lay out all  potential permutations and then calculates a p-value for us, and uses this distribution to come up with a measure. Lets use it to test the null hypothesis, using the `fisher.test()` function.   


```{r fisher, exercise=TRUE, eval = FALSE}
spider_wide %>%
    ___(-__)%>%
    fisher.test()
```

```{r fisher-solution}
spider_wide %>%
   select(-secondMaleAcceptance)%>%
   fisher.test()
```

Look at the results... Now pipe them into the `tidy()` function in broom


```{r fisher2, exercise=TRUE, eval = FALSE}
library(broom) 
spider_wide %>%
    ___(-__)%>%
    fisher.test() %>%
    ___()
```

```{r fisher2-solution}
spider_wide %>%
   select(-secondMaleAcceptance)%>%
   fisher.test() %>%
   tidy()
```



```{r conclude, echo = FALSE}
     question("Assuming a traditional alpha threshold of 0.05, what  do you do to the null hypothesis",
               answer("Reject it", correct = TRUE),
               answer("Fail to reject it"),
              allow_retry = TRUE,   
              correct = random_praise(),
              incorrect = random_encouragement())
```


## Chi2

An alternative method to null hypothesis significance testing would be to calculate $\chi^2$ and look this up on the $\chi^2$ sampling distribution.   Modify the code below to find expected counts in each category, assuming indepenence, so we can calculate $\chi^2$  

```{r expectations, exercise=TRUE, eval = FALSE}
spider_long %>%
  mutate(total_n = sum(count)) %>%
  group_by(secondMaleAcceptance)     %>% 
    mutate(p_accept = __ / __) %>% 
    ungroup() %>%
  group_by(eatenOrEscape) %>% 
    mutate(p_eatenOrEscape = __ / __) %>% 
    ungroup() %>%
  mutate(expected_count = __ * __ * __)
```


```{r expectations-solution}
spider_long %>%
  mutate(total_n = sum(count)) %>%
  group_by(secondMaleAcceptance)     %>% 
    mutate(p_accept = sum(count) / total_n) %>% 
    ungroup() %>%
  group_by(eatenOrEscape) %>% 
    mutate(p_eatenOrEscape = sum(count) / total_n) %>% 
    ungroup() %>%
  mutate(expected_count = total_n * p_accept * p_eatenOrEscape)
```



```{r expected, echo = FALSE}
     question_text("Under independence how many spiders do you expect would have been eaten and had their mates reject the second suitor? (round to the nearest integer)",
               answer("2", correct = TRUE),
              allow_retry = TRUE,   
              correct = random_praise(),
              incorrect = random_encouragement())
```

```{r expected2, echo = FALSE}
     question("What assumption of the chi2 test is broken, making the Fisher test more appropriate?",
               answer("Data are not independent"),
               answer("One of the categories has an expected count less than one"),
               answer("More than 20% e of the categories has an expected count less than five", correct = TRUE),
               answer("Data are biased"),
              allow_retry = TRUE,   
              correct = random_praise(),
              incorrect = random_encouragement())
```

Calculate $\chi^2$ anyway   

```{r chi2calc, message=FALSE, warning=FALSE, echo=TRUE, eval = FALSE}
spider_long %>%
  mutate(total_n = sum(count)) %>%
  group_by(secondMaleAcceptance)     %>% 
    mutate(p_accept = __ / __) %>% 
    ungroup() %>%
  group_by(eatenOrEscape) %>% 
    mutate(p_eatenOrEscape = __ / __) %>% 
    ungroup() %>%
  mutate(expected_count = __ * __ * __,
         each_chi2       = (___ - __)^2 / ___)%>%
  summarise(chi2 = __(__))
```

```{r chi2calc-solution}
spider_long %>%
  mutate(total_n = sum(count)) %>%
  group_by(secondMaleAcceptance)     %>% 
    mutate(p_accept = sum(count) / total_n) %>% 
    ungroup() %>%
  group_by(eatenOrEscape) %>% 
    mutate(p_eatenOrEscape = sum(count) / total_n) %>% 
    ungroup() %>%
  mutate(expected_count = total_n * p_accept * p_eatenOrEscape,
         each_chi2      = (expected_count - count)^2 / expected_count)%>%
  summarise(chi2 =sum(each_chi2))
```


```{r chi2val, echo = FALSE}
     question_text("What did you find chi2 to equal? (include one digit past the decimal)",
               answer("14.7", correct = TRUE),
              allow_retry = TRUE,   
              correct = random_praise(),
              incorrect = random_encouragement())
```

We can estimate a p-value by comparing our value of $\chi^2$ to the $\chi^2$ sampling distribution with one degree of freedom (2 columns - 1) x (2 rows - 1) = 1. As follows


```{r chi2p}
# Execute the code to find a p-value
pchisq(q = 14.7, df = 1, lower.tail = FALSE)
```

**Consider, why did we only look at the lower tail?** 


Run the code below to have R do a $\chi^2$ test for us  

```{r chi2R, message=FALSE, warning=FALSE, echo=TRUE, eval = FALSE}
spider_wide %>%
   select(-secondMaleAcceptance)%>%
   chisq.test(orrect = FALSE) %>% # tells R to calculate a standard chi2
   tidy()
```



```{r q, echo = FALSE}
     question("Compared to our chi2 math, Rs answer for the p-value and chi2 value was",
               answer("Basically the same", correct = TRUE),
               answer("Off by  a bit, but doesn't really change the story"),
               answer("Off by  a lot, be afraid"),
              allow_retry = TRUE,   
              correct = random_praise(),
              incorrect = random_encouragement())
```


```{r q2, echo = FALSE}
     question("Compared to the chi2 math, the p-value from Fisher's exact test was",
               answer("Basically the same"),
               answer("Off by  a bit, but doesn't really change the story", correct = TRUE),
               answer("Off by  a lot, be afraid"),
              allow_retry = TRUE,   
              correct = random_praise(),
              incorrect = random_encouragement())
```
