---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(knitr)
library(kableExtra)
library(ggmosaic)
library(cowplot)
library(infer)
library(broom)
knitr::opts_chunk$set(echo = FALSE)

```

# Normal quiz  


## Worked example  



The weights of singleton (that is, not twins) babies born in the U.S. are  (roughly) normally distributed with a mean of 3.339 kg with a standard deviation of 0.573 kg.  


#### How many standard deviations away from the mean is a 5 kg (11 lb) baby?  

```{r math, exercise=TRUE, eval = FALSE}
# mean = 3.339, sd  = 0.573
```

<div id="math-hint">
**Hint:** Divide the difference between the observation and the parameter by the standard deviation.
</div>



```{r z1, echo = FALSE}
question_text("How many  standard deviations away from the mean is a 5 kg (11 lb) baby?  (include three numbers past the decimal).",
 answer("2.898", correct = TRUE),
 answer("2.899", correct = TRUE),
 allow_retry = TRUE,   
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

```{r z2, echo = FALSE}
question("What is the name of the procedure you did to find the answer above.",
 answer("The Central Limit Theorem"),
 answer("The q transform"),
 answer("The Z transform", correct = TRUE),
 answer("The log transform"),
 allow_retry = TRUE,   
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

#### What is the probability density that a baby is 5kg? 


Use the appropriate [`_norm()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Normal.html) function to find the probability density that a baby is 5 kg (11 pounds)

```{r math11, exercise=TRUE, eval = FALSE}
# mean = 3.339, sd  = 0.573, weight = 5

```

<div id="math11-hint">
**Hint:** Its `dnorm()`.
</div>


```{r z21, echo = FALSE}
question("The answer above describes the proprtion of babies than will 5 kg at birth (be prepared to explain).",
 answer("TRUE"),
 answer("FALSE", correct = TRUE),
 allow_retry = TRUE,   
  correct = random_praise(),
  incorrect = random_encouragement()
)
```


Use the appropriate [`_norm()`](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Normal.html) function to find the probability  that a baby is greater than 5 kg (11 pounds)   

```{r math21, exercise=TRUE, eval = FALSE}
# mean = 3.339, sd  = 0.573, weight = 5

```

<div id="math21-hint">
**Hint:** Its `pnorm()`, and consider if `lower.tail` should be `TRUE` or `FALSE`. 
</div>


#### Exceptional babies   

What value separates the heaviest 99% of babies from the rest (hint, this is a one-tailed question, unlike examples in the book)


```{r math12, exercise=TRUE, eval = FALSE}
# mean = 3.339, sd  = 0.573, quantile = .99
 
```

<div id="math12-hint">
**Hint:** Its `qnorm()`, and consider if `lower.tail` should be `TRUE` or `FALSE`. 
</div>


```{r z12, echo = FALSE}
question_text("What value (in kg) separates the heaviest 99% of babies from the rest ?  (include three numbers past the decimal).",
 answer("4.672", correct = TRUE),
 allow_retry = TRUE,   
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

#### Probability of a range   

What is the probability a baby is between three and four kg?   


```{r math121, exercise=TRUE, eval = FALSE}
# mean = 3.339, sd  = 0.573, range 3 to 4

```


<div id="math121-hint">
**Hint:** Subtract the probability it is less than three from the probability it is less than four (use `pnorm()` for each). 
</div>


```{r z121, echo = FALSE}
question_text("What is the probability a baby is between three and four kg?  (include three numbers past the decimal).",
 answer("0.598", correct = TRUE),
 answer("0.599", correct = TRUE),
 answer(".598", correct = TRUE),
 answer(".599", correct = TRUE),
 allow_retry = TRUE,   
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

#### Baby sampling distributions   

Imagine we took five babies born in a given hospital and calculated their mean weight. If you repeated this again for all the hospitals in the country, you would have a sampling distribution, and if each sample was independent (hint it's not),  the standard deviation of this distribution would equal   

```{r math2, exercise=TRUE, eval = FALSE}
# mean = 3.339, sd  = 0.573, n = 5

```


```{r z3, echo = FALSE}
question_text("The standard deviation of the sampling distribution of the weights of five babies equals (include three numbers past the decimal).",
 answer("0.256", correct = TRUE),
 answer(".256", correct = TRUE),
 allow_retry = TRUE,   
  correct = random_praise(),
  incorrect = random_encouragement()
)
```



<div id="mathz3-hint">
**Hint:** Maybe the next question is all the hint you need?    
</div>

```{r z4, echo = FALSE}
question("We call this value the",
 answer("Standard deviation"),
 answer("Standard error", correct = TRUE),
 answer("Best linear unbiased predictor"),
 answer("Confidence interval"),
 answer("Z distribution"),
 answer("t distribution"),
 allow_retry = TRUE,   
  correct = random_praise(),
  incorrect = random_encouragement()
)
```

#### Simulate babies

Focus on one of the questions above, Use the appropriate `_norm()` function to simulate babies and convince yourself that the math is right.  


```{r math3, exercise=TRUE, eval = FALSE}
n_reps       <- 100000
sample_size  <- 5
baby_sd      <- 0.573
baby_mean    <- 3.339

tibble(hospital  = rep(1:n_reps, each = sample_size),
       weights   = _norm(n = n_reps * sample_size, mean = __, sd = ___)) %>%
  __(__) %>%
  summarise(mean_vals = mean(weights))%>%
  summarise(se = __(___))
```


```{r math3-solution, exercise=TRUE, eval = FALSE}
n_reps       <- 100000
sample_size  <- 5
baby_sd      <- 0.573
baby_mean    <- 3.339

tibble(hospital  = rep(1:n_reps, each = sample_size),
       weights   = rnorm(n = n_reps * sample_size, mean = baby_mean, sd = baby_sd)) %>%
  group_by(hospital) %>%
  summarise(mean_vals = mean(weights))%>%
  summarise(se = sd(mean_vals))
```



#### Think  

Do you think baby weights  are actually perfectly normal? If not, how will they deviate from the normal distribution?  

##  Normality   

```{r z5, echo = FALSE}
question("The normal distribution (or something pretty close to it) is quite ____ in the real world",
 answer("Rare"),
 answer("Common", correct = TRUE),
 allow_retry = TRUE,   
  correct = random_praise(),
  incorrect = random_encouragement()
)
```


```{r z6, echo = FALSE}
question("The central limit theorem states that all sampling distributions will be normal, regardless of sample size or the underlying distribution.",
 answer("TRUE"),
 answer("FALSE", correct = TRUE),
 allow_retry = TRUE,   
  correct = random_praise(),
  incorrect = "Incorrect. What sample size do we need for the central limit theorem?"
)
```


Recall or app to visually determine the sample size required to see  if sampling distributions are roughly normally distributed  (ie how big of a sample we need to have normally distributed sampling distributions)

```{r}
include_app("https://brandvain.shinyapps.io/centrallimit/",height = "750")
```


**Rank the distributions above, from which needs the smallest to largest sample size to have a normal sampling distribution**

#### Evaluating normality and data transformation

```{r}
my_dat <- tibble( x = (rpois(75, lambda  = 2.2) )) 
        a <- ggplot(my_dat , aes(x = x)) + geom_histogram(binwidth =  1,color = "white")+labs(title = "Histogram")
        b <- ggplot(my_dat , aes(x = x)) + geom_density(fill = "grey")+labs(title = "Density plot")
        c <- ggplot(my_dat , aes(sample = x)) + geom_qq() + geom_qq_line()+labs(title = "quantile-quantile plot")
        d <- ggplot(my_dat , aes(x = x)) + stat_ecdf()+labs(title = "Cumulative distribution")
        plot_grid(plot_grid(a,b,c, d, ncol = 2))
```


```{r z7, echo = FALSE}
quiz(caption = "Consider the plot above.", 
  question("What is a giveaway its not normal (pick the best answer).",
 answer("The mean isn't zero"),
 answer("Data are not symmetric about the mean", correct = TRUE),
 answer("More than 95% of the data is two standard deviations away from the mean"),
 allow_retry = TRUE,   
  correct = random_praise(),
  incorrect = random_encouragement()), 
  question("What transformation is most appropriate to make it normal? ",
 answer("A Z-transform"),
 answer("An arcsin square  root transform"), 
 answer("An reciprocal transform"),
 answer("A log(x) transform"),
 answer("A log(x+2) transform", correct = TRUE),
 answer("No transform is needed"),
 allow_retry = TRUE,   
  correct = random_praise(),
  incorrect = random_encouragement()
))
```


